---
title: "Intro to Raster"
author: "Loïc Dutrieux, Jan Verbesselt"
date: "01/06/2015"
output: ioslides_presentation
---

<--
Include:
Reminder on raster classes
Add, substract raster layers
simple value replacement
calc and overlay
-->


# Introduction

**Today**

Introduction to raster manipulation and analysis


    * Read and write raster data
    * Raster arithmetic
    * Raster manipulation: re-projection, format conversion, cropping, layer stacking, etc
    * Apply function to single and multiple raster objects


**Tomorrow**

Advanced raster analysis (Filtering, classification, etc...)

**Next week**

Temporal analysis of raster data

---

# What's under the hood
## Libraries, bindings, packages

**GDAL**: Geospatial Data Abstraction Library

    * Contains the drivers for reading all formats of data


**GDAL** and **PROJ4**

    * Performs geographical transformations of raster data.


The `rgdal` package is the interface between R and these libraries.


---

# The raster package
Features:

    * Read and write raster data (Makes extensive use of the `rgdal` package)
    * Perform most raster operations (create raster objects, perform spatial/geometric operations (re-projections, resampling, etc), filtering and raster calculations)
  * Can work on big raster objects
  * Relatively fast
  * Visualize and interact with the data
  * Extremely well documented (check the vignettes)
  * etc...


Load the raster package using the following command.
```{r, eval=FALSE}
library(raster)
```

---

# The raster objects
3 main classes of raster object exist in the raster package.

    * RasterLayer: Single georeferenced raster object
    * RasterBrick: Multilayer georeferenced raster object
    * RasterStack: Virtual RasterBrick


```{r, eval=FALSE}
# Create a raster object and add random values to it
r <- raster(nrow=20, ncol=40)
r[] <- rnorm(n=ncell(r))
r
# Stack the layer 3 times
s <- stack(r,r,r * 2)
s
# Coerce to RasterBrick object
b <- brick(s)
b
```


```{r, echo=FALSE, results='asis'}
library(knitr)
df <- data.frame(Class = c('RasterLayer', 'RasterBrick', 'RasterStack'), Description = c('Single layer raster object',  'Multiple Layer Raster object', 'Multiple Layer Raster object'))
kable(df)
```

---

# Reading data
For single layer raster objects, simply use the `raster()` command.

See the help page of the function.
```{r, eval=FALSE}
?raster
```

For multilayer objects, use `stack()` or `brick()`.



Example: Read, visualize the Landsat data stored in the extdata/ folder of the rasta package.
```{r, eval=FALSE}
# Get the path to the data
x <- system.file('extdata/LE71700552001036SGS00_SR_Gewata_INT1U.tif',
                 package= 'rasta')
gewata <- brick(x)
```


Now, let's write this object to the 'D:/' drive.
```{r, eval=FALSE}
writeRaster(x=gewata, filename='D:/filepath/filename.tif')
```

---

# Simple Raster calculations
Raster arithmetic follows simple R vector arithmetic rules.
Therefore to perform pixel based calculations, simply use the operators +, -, *, /



$NDVI=\frac{NIR-R}{NIR+R}=\frac{band 4 - band 3}{band 4 + band 3}$


```{r, eval=FALSE}
ndvi <- (gewata[[4]] - gewata[[3]]) / (gewata[[4]] + gewata[[3]])
plot(ndvi)
```

These types of operations are not advised for big rasters as they can overload the R internal memory.

Instead, `calc()` and `overlay()` should be used.

---

# calc() and overlay()


    * Both function implement block processing (no memory limitation)
    * `calc()` mostly for applying a function to single raster objects
    * `overlay()` makes it easier to implement function that involve multiple raster objects
    * These functions are only suitable for pixel based calculation


\centerline{**overlay example}**
```{r, eval=FALSE}
# That function has two input arguments, therefore it is suitable
# for overlay()
ndviOver <- function(red, nir) {
    out <- (nir - red) / (nir + red)
    return(out)
}

ndvi <- overlay(x=gewata[[3]], y=gewata[[4]], fun=ndviOver)
```


---

# calc() and overlay() (2)


\centerline{**calc example}**
```{r, eval=FALSE}
ndviCalc <- function(x) {
    out <- (x[[4]] - x[[3]]) / (x[[3]] + x[[4]])
    return(out)
}

ndvi2 <- calc(x=gewata, fun=ndviCalc)
```

Another advantage of `calc()` and `overlay()` is that they can perform calculations directly from file to file, see the **filename=** argument hidden in the ellipsis.

```{r, eval=FALSE}
calc(x=gewata, fun=ndviCalc, filename='filepath/filename.tif')
```
The **filename=** argument holds for most of the functions of the raster package.


---

# Raster objects manipulation
But where has this gewata scene beeen acquired by the way?

To discover that, we can try to project the NDVI layer in Google Earth.

```{r, eval=FALSE}
ndviLL <- projectRaster(ndvi, crs='+proj=longlat')
KML(x=ndviLL, filename='filepath/filename.kml')
```
The entire database of projections is at [http://spatialreference.org/]

**Another type of raster object manipulation.**

    * Crop data interactively.



```{r, eval=FALSE}
plotRGB(gewata, 4,5,3, stretch= 'lin')
e <- drawExtent()
gewata2 <- crop(x=gewata, y=e)
plotRGB(gewata2, 4,5,3, stretch= 'lin')
```


---

# Other types of raster object manipulations
## Layer stacking

Now let's prepare the data for tomorrow's lecture.

    * [http://rasta.r-forge.r-project.org/]
    * We want to:
    
    
        * Read the data
        * Create a layer stack object on disk
    



```{r, eval=FALSE}
getwd()
# Downlaod and unpack data
download.file(url='http://rasta.r-forge.r-project.org/tura.zip',
              destfile='tura.zip')
unzip(zipfile='tura.zip')
# Retrieve the content of the tura sub-directory
list <- list.files(path='tura/', full.names=TRUE)
# Quickly visualize the first layer
plot(raster(list[1]))
# Stack the layers
turaStack <- stack(list)
turaStack
# Write to disk
writeRaster(x=turaStack, filename='filename.grd', datatype='INT2S')
```

---

# Reminder about storage and data types
Check ?dataType

{c  c  c}
**Datatype definition** & **minimum possible value** & **maximum
possible value** \cr
`INT1S` & -127 &  127 \cr
`INT1U` & 0 &  255 \cr
`INT2S` & -32,767 &  32,767 \cr
`INT2U` & 0 &  65,534 \cr
`INT4S` & -2,147,483,647 & 2,147,483,647 \cr
`INT4U` & 0 & 4,294,967,296 \cr
`FLT4S` & -3.4e+38 &  3.4e+38 \cr
`FLT8S` & -1.7e+308 & 1.7e+308 \cr


---

# Summary so far
So far we saw how to:

    * Read and write raster data
    * perform simple raster algebra
    * use calc() and overlay()
    * crop
    * re-project
    * stack multiple layers


---

# More advanced raster arithmetic
## perform value replacements
Remember vector arithmetic?
```{r, eval=FALSE}
a <- c(1,2,3,4,2,3,7)
a <- a * 2
a[a == 4] <- NA
b <- c(1,0,0,1,0,0,0)
a[b == 1] <- NA
```

Data (pixel values) are stored in Raster objects in the form of vectors. Therefore we can use vector arithmetic to perform raster arithmetic.

Example: We can replace all the values <0.4 in the first layer of the gewata NDVI layer by NAs.
```{r, eval=FALSE}
ndvi[ndvi < 0.4] <- NA
plot(ndvi)
```

---

# More about replacements
The rasta package contains a built in dataset called taravao.
```{r, eval=FALSE}
library(rasta)
data(taravao)
```

    * It is a subset of a Landat scene
    * It covers part of the Taravao village, on Tahiti island
    * It contains a QA (Quality assessment) band in Layer 9
    
        * Contains information on the quality of the data
        * The **QA2cloud\{rasta\}** function can generate a cloud mask from this band
    


**Tutorial**


    * Visualize the data
    * Genarate the cloud mask
    * Drop band 9 from the rasterBrick
    * Replace cloudy pixels by NAs

 