---
title: "Intro to Raster"
author: "Loïc Dutrieux, Jan Verbesselt"
date: "01/08/2015"
output: slidy_presentation
---


# Yesterday

Project including function to check whether an input year is a leap year or not.

<center>
<h2>Who submitted</h2>
</center>

```{r, echo=FALSE, results = 'asis'}
suppressPackageStartupMessages(library(googleVis))
x <- data.frame(status = "submitted", n = 13)
pie <- gvisPieChart(data = x)
print('<center>')
print(pie)
print('</center>')
```

---

Well structured projects

`'error: argument of class numeric expected'` instead of `error()`

```{r, eval = FALSE}
rprint("No Leap Year")
## instead of
return()
## at the end of a function.
```

```{r, eval=FALSE}
return(year%%4==0)
```

---

# Today

Morning

* Go through [tutorial](http://geoscripting-wur.github.io/IntroToRaster/)

Afternoon

* Quick overview of the tutorial
* Exercise


---

# The raster objects (classes)


```{r, echo=FALSE, results='asis'}
library(knitr)
df <- data.frame(Class = c('RasterLayer', 'RasterBrick', 'RasterStack'), Description = c('Single layer raster object',  'Multiple Layer Raster object', 'Multiple Layer Raster object'))
kable(df)
```





* `raster()`
* `brick()`
* `stack()`
* `writeRaster()`

---

# Simple Raster calculations / raster arithmetic

Raster arithmetic follows simple R vector arithmetic rules.

* Value replacement
* Additions, substractions, etc

**Although warning, not memory safe for big rasters:**

* Instead, `calc()` and `overlay()` should be used.

---

# calc() and overlay()


* Both function implement block processing (no memory limitation)
* `calc()` mostly for applying a function to single raster objects
* `overlay()` makes it easier to implement function that involve multiple raster objects
* These functions are only suitable for pixel based calculation
* Can write output directly to disk (see `filename =` argument)

---

# Visualize raster data

* `plot()`
* `plotRGB()`
* On google Earth by converting to KML format or using `plotKML` package

Also possible using `ggplot2` and `rasterVis` package


 